A Framework for Discrete Change

### 2008-12-08 papers/individual/individual.tex
Finished individual paper, submitted version
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@247 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\shorttitle{Modelling Discrete Change}

@@

% intro
Discrete change frequently occurs in learning and development: in
learning concepts and discrimination learning, in performance on
Piagetian tasks such as the conservation of liquid task and the
balance scale task, and in conditioning.  This chapter is concerned
with detecting and characterizing such discrete changes in
(individual) time series data.  We present a framework of dependent
mixture models that can be used to test whether discrete learning
events are present and, if so, to test at which point in time those
changes are taking place.  Before presenting the model in formal terms
and providing some illustrations, we first review some examples from
the developmental psychology literature in which discrete change is
typically found.

@@

% hysteresis on the balance scale
\citet{Jansen2001b} found evidence for stage-wise transitions, or
discrete change in our model, between Rule 1 and Rule 2 by testing
criteria that were derived from the catastrophe model.  In particular,
they found bimodal test scores and inaccessibility.  The latter means
that there are no in-between strategies: children apply {\em either}
Rule 1 {\em or} Rule 2 and there is no in-between option.
\citet{Jansen2001b} also found evidence for hysteresis: the phenomenon
that switching between strategies is asymmetric.  Children can switch
from Rule 1 to Rule 2 and back, but this occurs at different values of
a continuously changing independent variable.  In particular, if the
distance dimension in the balance scale problems is made increasingly
more salient by increasing the distance difference between weights on
either side of the balance scale, children may switch from Rule 1 to
Rule 2.  If subsequently the distance difference is decreased again,
children may switch back to using Rule 1.  Hysteresis is the
phenomenon that this switch back occurs at a different value of the
control variable, in this case the distance difference.

@@

% discrimination and categorization learning
Sudden transitions in learning are also observed in simple
discrimination learning paradigms in which participants learn to
discriminate a number of stimuli based on a single dimension such as
form or color.  \citet{Raijmakers2001} found evidence for two
different strategies applied by children when faced with such a
learning task.  \citet{Schmittmann2006} reanalyzed their data using
hidden Markov models to show that both strategies are characterized by
discrete changes in the learning process.

% latent Markov versus time series analyses/outline
In above mentioned applications, the data consist mostly of a few
repeated measurements administered to large groups of participants.
The focus in the current chapter is rather on data that consist of
many repeated measurements, or time series, observed in only a few
participants or even just a single participant.  For example,
\citet{Visser2009c} analyzed data from a single participant in an
experimental task that manipulates the trade-off between speed and
accuracy.  The data consisted of three time series with each around
150 repeated measurements of both reaction time and accuracy.  Below
we provide examples of analyzing time series from single participants
from two experiments; one from the Iowa Gambling Task and one from the
Weather Prediction Task.  The interest in these tasks is to show that
participants develop different strategies over time in responding to
the stimuli and that the transition from one strategy to the next is a
discrete event.  Before providing these illustrations, we give a
formalization of dependent mixture models and a brief overview of the
\pkg{depmixS4} package that was developed to specify and fit such
models.

@@

% outline
In this section we describe a class of models which are especially
suitable for describing and testing discrete changes in (individual)
time series data.  The dependent mixture model is similar to, but
slightly different from, two other types of models that are in use for
modelling discrete change: the hidden and the latent Markov models.

% history of Markov models/LMM applications
Markov models have been used for a long time in the social sciences;
for example, in analyzing language learning
\cite{Miller1952,Miller1963} and in the analysis of paired associate
learning \cite{Wickens1982}.  In these models, the focus is on survey
type data: a few repeated measurements taken in a large sample.
\citeA{Langeheine1990} discuss latent Markov models and their use in
sociology and political science \citep<see also>{McCutcheon1987}.  Latent
Markov models have been used in studying development of math skills
\citep{Collins1992}, and in medical applications
\citep{Reboussin1998}.  \citet{Kaplan2008} provides an overview of
such models, also called stage-sequential models, and their
application in developmental psychology.

% hidden Markov models
Hidden Markov models (HMMs) tend to be used in the analysis of long
univariate (individual) timeseries.  For example, HMMs are the model
of choice in speech recognition applications \cite{Rabiner1989}.  In
biology, HMMs are used to analyze DNA sequences \cite{Krogh1998} and
in econometric science, to analyze changes in stock market prices and
commodities \cite{Kim1994}.

% depmix model
The dependent mixture model (DMM) that we propose here spans the range
from latent Markov models for few repeated measurements with many
participants to hidden Markov models for individual time series.  In
addition, the dependent mixture model includes multivariate responses
and covariates.  The dependent mixture model for a time series $O_{t}$
consists of the following elements:
\begin{enumerate}
	\item $S$ is a collection of discrete states, forming the state space.
	\item $S_{t} = \mat{A}S_{t-1}+\xi_{t}, \mat{A}$, a transition matrix.
	\item $O_{t} = \mat{B}(S_{t}) + \zeta_{t}, \mat{B}$,  an observation density.
\end{enumerate}
Here $\xi_{t}$ and $\zeta_{t}$ are independent error processes 
\cite{Elliott1995}.

@@

% observation densities
The observation densities $\mat{B}$ form the measurement part of the
model; these describe the distributions of the observations
conditional on the current state. Hence, these distributions
characterize the state, and in our examples, these characterize the
rule that a participant is using at a given measurement occasion.

@@

% outline/abstract: this has redundant information, mostly discussed
% in detail below
\pkg{depmixS4} \citep{Visser2008a} implements a general framework for
defining and fitting dependent mixture models in the R programming
language \citep{R2008}.  This includes standard Markov models,
latent/hidden Markov models, and latent class and finite mixture
distribution models.  The models can be fitted on mixed multivariate
data with error distriutions from the generalized linear model
framework, such as the gaussian, binomial, multinomial logistic et
cetera.

@@

@@

\pkg{depmixS4} is built using S4 classes (object oriented classes in R)
to allow easy extensibility (Chambers, 1998). A depmix model consists 
of the following: 
\begin{enumerate} 
	\item A model for the initial or prior probabilities.
	\item A list of transition models; one model for each row of the 
	transition matrix.
	\item A list of response models for each of the states of the model; 
	this is a list of lists in the case of multivariate responses. 
\end{enumerate} 
Each of these are adressed briefly below.

@@

% transition and initial state probs
By default, each row of the transition matrix and the vector of
initial state probabilities is modelled as a baseline-category
multinomial logistic model \cite<see>[chapter 7]{Agresti2002}.  This
means that covariates can be used as predictors in these models.  In
particular, the model for each multinomial is:
\begin{equation} 
	\pi_{i} = \frac{\exp(\alpha_{i}+\mathbf{\beta}^{`}_{i}\vc{x})}
				{1+\sum_{j=1}^{J-1}\exp(\alpha_{j}+\mathbf{\beta}^{`}_{j}\vc{x})},
\end{equation}
where $\pi_{i}$ is the probability (e.g.\ the probability of the
$i$-th initial state); the $\alpha_{i}$ are the category intercepts;
the $\mathbf{\beta}_{i}$ are the category regression coefficients;
$\vc{x}$ is a vector of covariates or predictors; $J$ is the number of
states; in this example, $J$ serves as the baseline-category, meaning
that $\alpha_{J}$ and $\mathbf{\beta}_{J}$ are zero.

@@

% response models
Response models in \pkg{depmixS4} interface with the \code{glm}
functions \citep{R2008} available in R and with the \pkg{nnet}-package
\citep{Venables2002} for the multinomial logistic models.  Normally
distributed data can hence be modelled with direct effects included as
well.  For example, normal data are modelled as:
\begin{equation}
	O_{t}|S_{t} = \mu + \mathbf{\beta}^{`}\vc{x} + \epsilon_{t},
\end{equation}
where $\mu$ is the (state-dependent) mean; $\vc{x}$ is a vector of
covariates or predictors; $\mathbf{\beta}$ is the vector of
(state-dependent) regression coefficients; $\epsilon_{t}$ is a
normally distributed error term with (state-dependent) standard
deviation $\sigma$.

@@

Two illustrations are provided below of models that analyze single
participant time series data from two common experimental paradigms.
In both of these, participants learn different strategies through
trial and error.  The main goal of these illustrations is to establish
the possibility of analyzing single participant data using the
\pkg{depmixS4}-framework.

@@

% strategic reanalysis of the HDT: different strategies found by
% finite mixture analysis
A reanalysis of this dataset \cite{Huizenga2007b} indicated that
participants might solve the task by sequentially considering the
three dimensions (constant gain, frequency of loss, and amount of
loss) in order to choose a door.  Most of the youngest children seem
to focus on frequency of loss, resulting in equal preference for doors
B and D. It is assumed that frequency of loss is the dominant
dimension in this task.  Older participants seem to use a
two-dimensional rule where participants first focus on the frequency
of loss and then consider amount of loss, resulting in a preference
for door D. A third very small subgroup seems to use an integrative
rule where participants combine all three dimensions in the
appropriate way.  Participants using the integrative rule pick cards
from doors C and D, which are both equally advantageous in the long
run, although door D shows a higher volatility.

@@

% problematic aspects of standard analysis
\citet{Huizenga2007b} used the last 60 trials in a series of 200
trials, as is typically done in analyzing this task.  Doing so
silently assumes that behavior has stabilized after 140 trials of
learning; this may not be the case, and it is highly likely that there
are individual differences in this learning process.  Hence, here we
analyze IGT learning data from a single participant with the aim to
establish 1) at which trial learning stops, i.e., at which trial
behavior has stabilized, and 2) what strategies participants use
during and at the end of learning.  Models with an increasing number
(from 2 through 5) of latent states were fitted to the time series of
a single participant.  The responses were fitted with multinomial
logistic models for the four possible choices that participants make
in this task.  There were no covariates on any of the parameters.  The
only constraint imposed on the models' parameters was that there were
designated begin and end states.  This means that the initial state
probabilities were fixed to one for the first state and to zero for
the remaining states.  For the transition matrix this means that the
final state was an absorbing state: the probability of transitioning
out of the final state was set to zero.  This was done to ensure that
there would be a final state in which participants end which provides
the possibility of immediately seeing participants' behavior at the
end of the task.  Models were selected using the Akaike Information
Criterion \citep<AIC;>{Akaike1973}.

The data were best described by a 4-state model.  This model's
transition matrix is:
$$
\mat{A} = \begin{pmatrix} 
				0.64 & 0.33 & 0.00 & 0.03 \\
				0.10 & 0.89 & 0.01 & 0.00 \\
				0.05 & 0.00 & 0.95 & 0.00 \\
				0.00 & 0.00 & 0.00 & 1.00
		  \end{pmatrix}
$$
As can be seen from the transition matrix, the final state is
absorbing.  States 2 and 3 are also fairly stable with high diagonal
probabilities of 0.89 and 0.95 respectively.  The response parameters
are presented in Table~\ref{tab:igt4} below.

@@

The initial preference for B and D choices confirms the theory
expressed in \citet{Huizenga2007b} that frequency of loss is the
dominant dimension in the IGT. The final states with mostly C 
choices represents one of the optimal strategies, which consists of
both C and D responses. Note that C and D responses generate equal
profits in the long run.

@@

There are different accounts of probabilistic category learning
\citep<see e.g.>[for an overview]{Ashby2005}.  According to instance
or exemplar learning theories, participants learn by storing each
encountered cue-outcome pair.  When presented with a probe cue
pattern, a response is made by retrieving these exemplars from memory
and weighting them according to their similarity to the probe cue
pattern.  In contrast, according to associative learning theories,
participants learn by gradually associating the individual cues (or
cue patterns in configural learning) to the outcomes.  Finally, in
rule-learning, participants are taken to extract rules by which to
categorize the different cue patterns.  \citet{Gluck2002} proposed a
number of such rules (or strategies).  The rules differ in the way the
cues are combined into a response.  The main difference concerns
whether responses are based on the presence/absence of a single cue,
or whether information from multiple cues is integrated.
\citet{Gluck2002} formulated all strategies in a deterministic and
optimal manner (e.g., the multi-cue strategy corresponded to giving
the optimal response to each cue pattern).  \citet{Meeter2006} allowed
for probabilistic responding (a small probability of responding
non-optimally) as well as switches between strategies during learning.
However, both strategies and response probabilities remained
predefined.  Alternative non-strategy based analyses of the WPT
\cite{Lagnado2006,Speekenbrink2008} have estimated response
distributions through logistic regression, allowing the regression
coefficients to change smoothly over time.

%associative, rule-based.
Here, we combine the regression and strategy-based approaches, and
analyse the behavior of a single individual performing the WPT for 200
trials.  We are particularly interested in evidence for strategy
switching and whether a DMM can recover strategies in accordance with
\citet{Gluck2002}.  We chose to analyse the ``average'' participant
(the participant with performance closest to the group average) in a
large unpublished dataset.  We let each state be characterized by a
GLM with a Binomial distributed response and logistic link function
(i.e., a logistic regression model, see equation 1).  The regression
coefficients of a model relating the cues to the state of the weather
are given in column ```validity'' of Table~\ref{tab:WPT}.  Note that
identical coefficients for the model relating the cues to responses
would indicate probability matching.  A maximizing strategy is
indicated by more extreme regression coefficients.

@@

Investigation of the parameter estimates (see Table~\ref{tab:WPT})
showed that the regresion coefficient for the first cue was of much
larger magnitude than that of the other cues.  Because of this, the
first state seems representative of single cue strategy.
Alternatively, as all regression coefficients are positive, it could
indicate a ``counting'' heuristic, where the propensity of ``sun''
responses increases when more cues are present (regardless of which
cues they are).  However, in that case, we would expect the regression
coefficients to be of roughly identical magnitude.  The second state
represents a multi-cue strategy, as all cues had regression
coefficients of reasonable magnitude in the direction of the objective
cue validities.

To reduce the degrees of freedom and improve parameter estimates, we
implemented constraints to force state 1 into a single cue strategy
(fixing the coefficients of the remaining three cues to 0) and state 2
in a multi-cue strategy (forcing the intercept to 0).  These
restrictions resulted in a better AIC value of AIC=185.24 (df=7).
Interestingly, the single cue strategy was somewhat different than
described by \citet{Gluck2002}.  Parameter estimates indicated relatively
more consistent predictions of ``rain'' in the absence of cue 1
($Pr(\text{sun}) = 0.22$) and more inconsistent predictions of ``sun''
in the presence of cue 1 ($Pr(\text{sun}) = 0.60$).  The cue weights
of the multi-cue strategy were in the direction of the optimal
weights.  The Viterbi state sequence indicated that the participant
used the single cue strategy for the first 60 trials, and then
switched to the multi-cue strategy.

@@

Discrete change is hypothesized to occur in many domains in
experimental psychology and especially so in learning and
developmental tasks.  Piagetian theory predicts discrete changes to
occur in a number of developmental tasks \citep{Inhelder1958}.  There
is a growing body of evidence showing that indeed discrete change
occurs during development and learning
\citep{Jansen2001b,Maas1992,Raijmakers2001,Schmittmann2006}.  Discrete
strategies or rules have recently been hypothesized to play an
important role in learning tasks such as the Iowa Gambling Task and
the Weather Prediction Task.  Until now, such data were always
analyzed on a group basis, i.e., pooling data of many participants
together.  Given that it is extremely likely that there are individual
differences between participants, such pooling of data may not be
warranted and may lead to sub optimal results \citep{Molenaar2005}.

In the current study we have shown the feasibility of analyzing single
participant data in two common experimental tasks, thereby avoiding
the pitfalls of pooling data from individuals that may differ in
important respects.  The illustrations show that single participant
data only comprising 200 trials can be analyzed and lead to
non-trivial models with multiple states.  This opens up possiblities
for analyzing data from other experimental tasks on an individual
basis.

In addition to analyzing single participant data, the \pkg{depmixS4}
package \citep{Visser2008b} that we have presented here offers the
possibility of further combining models of single participant data
into larger models and testing whether for example some strategies or
rules are identical between (groups of) participants.  Performing such
analyses and aggregating models afterwards based on commonalities and
differences between participants avoids making unwarranted assumptions
about homogeneity between participants.  A similar approach using
continuous latent variable models, i.e. factor models, was used by
\citet{Hamaker2007} to integrate within and between person variability
in the domain of personality research.

% IGT
The illustration with the IGT data lead to a four state model.  The
states of the model corresponded very well with earlier analyses by
\citet{Huizenga2007b}, showing a number of discrete strategies.  Even
within a single participant, a number of different strategies were
observed during the acquisition process; learning started with
behavior close to guessing, which could be interpreted as exploring
behavior, and was followed by two states with close to optimal
performance.  In addition to confirming the existence of discrete
strategies, the use of DMM analyses provides us with information of
the dynamics of knowledge acquisition in this task.

% WPT
Analyzing the WPT data on an individual level allowed us to precisely
estimate idiographic strategies and their progression during learning.
Moreover, using DMMs, we can combine an individual and group level
analysis, increasing reliability of the individual analyses whilst
allowing for substantial individual differences.  Previous analyses
\cite<e.g.,>{Meeter2006} pre-defined individual strategies, based on
participants' self-reports in \citet{Gluck2002}.  Interestingly,
\citet{Gluck2002} noted that these self-reports often did not
correspond to the strategies evident in participants' responses.  It
is well known that people may find it difficult to verbalize the way
in which they integrate multiple sources of information.  As such,
estimating strategies directly from participants' responses will
result in more valid assessment of their response strategies, and
provides a direct test of the validity of the strategy set identified
by \citet{Gluck2002}.

Possible future extensions in depmixS4
\begin{enumerate}
	\item richer measurement models, eg factor models, AR models etc
	\item richer transition models, eg continuous time measurement occasions
	\item explicit state durations
	\item identifiability of models
	\item model selection
\end{enumerate}

General issues
\begin{itemize}
	\item illustrations show results that are consistent with usual group based analyses AND 
	provide extra information on switching between strategies over the course of learning
	\item Many applications in experimental psychology: examples????
\end{itemize}

>>>>

\shorttitle{Modeling Discrete Change}

@@

% intro
Discrete change frequently occurs in learning and development: in
learning concepts and discrimination learning, in performance on
Piagetian tasks such as the conservation of liquid task and the
balance scale task, and in conditioning.  This chapter is concerned
with detecting and characterizing such discrete changes in
(individual) time series data.  We present a framework of dependent
mixture models that can be used to test whether discrete learning
events are present and, if so, to test at which point in time those
changes are taking place.  We first review some examples from the
developmental psychology literature in which discrete change is
typically found, and then turn to describing the model more formally.

@@

% hysteresis on the balance scale
\citet{Jansen2001b} found evidence for stage-wise transitions, or
discrete change in our model, between Rule 1 and Rule 2 by testing
criteria that were derived from the catastrophe model.  In particular,
they found bimodal test scores and inaccessibility.  The latter means
that there are no in-between rules: children apply {\em either} Rule 1
{\em or} Rule 2 and there is no in-between option.
\citet{Jansen2001b} also found evidence for hysteresis: the phenomenon
that switching between rules is asymmetric.  Children can switch from
Rule 1 to Rule 2 and back, but this occurs at different values of a
continuously changing independent variable.  In particular, if the
distance dimension in the balance scale problems is made increasingly
more salient by increasing the distance difference between weights on
either side of the balance scale, children may switch from Rule 1 to
Rule 2.  If subsequently the distance difference is decreased again,
children may switch back to using Rule 1.  Hysteresis is the
phenomenon that this switch back occurs at a different value of the
control variable, in this case the distance difference.

@@

% discrimination and categorization learning
Sudden transitions in learning are also observed in simple
discrimination learning paradigms in which participants learn to
discriminate a number of stimuli based on a single dimension such as
form or color.  \citet{Raijmakers2001} found evidence for two
different rules applied by children when faced with such a
learning task.  \citet{Schmittmann2006} reanalyzed their data using
hidden Markov models to show that both rules are characterized by
discrete changes during the learning process.

% latent Markov versus time series analyses/outline
In above mentioned applications, the data consist mostly of a few
repeated measurements administered to large groups of participants.
The focus in the current chapter is rather on data that consist of
many repeated measurements, or time series, observed in only a few
participants or even just a single participant.  For example,
\citet{Visser2009c} analyzed data from a single participant in an
experimental task that manipulates the trade-off between speed and
accuracy.  The data consisted of three time series with each around
150 repeated measurements of both reaction time and accuracy.  Below
we provide examples of analyzing time series from single participants
from two experiments; one from the Iowa Gambling Task and one from the
Weather Prediction Task.  The interest in these tasks is to show that
participants develop different rules over time in responding to the
stimuli and that the transition from one rule to the next is a
discrete event.  Before presenting these illustrations, we provide a
brief formal definition of dependent mixture models and an overview of
the \pkg{depmixS4} package that was developed to specify and fit such
models.

@@

% outline
In this section we describe a class of models which are especially
suitable for describing and testing discrete change in (individual)
time series data.  The dependent mixture model is similar to, but
slightly different from, two other types of models that are in use for
modeling discrete change: the hidden and the latent Markov models.

% history of Markov models/LMM applications
Markov models have been used for a long time in the social sciences;
for example, in analyzing language learning
\cite{Miller1952,Miller1963} and in the analysis of paired associate
learning \cite{Wickens1982}.  \citeA{Langeheine1990} discuss latent
Markov models and their use in sociology and political science
\citep<see also>{McCutcheon1987}.  Latent Markov models have been used
in studying development of math skills \citep{Collins1992}, and in
medical applications \citep{Reboussin1998}.  \citet{Kaplan2008}
provides an overview of such models, also called stage-sequential
models, and their application in developmental psychology.

% hidden Markov models
Hidden Markov models (HMMs) tend to be used in the analysis of long
(univariate, individual) time series.  For example, HMMs are the model
of choice in speech recognition applications \cite{Rabiner1989}.  In
biology, HMMs are used to analyze DNA sequences \cite{Krogh1998} and
in econometric science, to analyze changes in stock market prices and
commodities \cite{Kim1994}.

% depmix model
The dependent mixture model (DMM) that we propose here spans the range
from latent Markov models for few repeated measurements with many
participants to hidden Markov models for individual time series.  In
addition, the dependent mixture model includes multivariate responses
and covariates.  The dependent mixture model for a time series $O_{t}$
consists of the following elements:
\begin{enumerate}
	\item $S$ is a collection of discrete states, forming the state space.
	\item $S_{t} = \mat{A}S_{t-1}+\xi_{t}, with \mat{A}$ a transition matrix.
	\item $O_{t} = \mat{B}(S_{t}) + \zeta_{t}, with \mat{B}$  an observation density.
\end{enumerate}
Here $\xi_{t}$ and $\zeta_{t}$ are independent error processes 
\cite{Elliott1995}.

@@

% observation densities
The observation densities $\mat{B}$ form the measurement part of the
model; these describe the distributions of the observations
conditional on the current state $S_{t}$. Hence, these distributions
characterize the state, and in our examples, these characterize the
rule that a participant is using at a given measurement occasion.

@@

% outline/abstract: this has redundant information, mostly discussed
% in detail below
The \pkg{depmixS4} package \citep{Visser2008a}, which is used to
analyze data in the illustrations below, implements a general
framework for defining and fitting dependent mixture models in the R
programming language \citep{R2008}.  This includes standard Markov
models, latent/hidden Markov models, and latent class and finite
mixture distribution models.  The models can be fitted on mixed
multivariate data with error distributions from the generalized linear
model framework, such as the gaussian, binomial, multinomial logistic
et cetera.

@@

	\item to be able to fit models on arbitrary length individual time 
	series.

@@

\pkg{depmixS4} is built using S4 classes \citep<object oriented
classes in R>{Chambers1998} to allow easy extensibility.  A depmix
model consists of the following:
\begin{enumerate} 
	\item A model for the initial or prior probabilities, possibly 
	depending on covariates.
	\item A list of transition models; one model for each row of the 
	transition matrix, possibly depending on covariates. 
	\item A list of response models for each of the states of the model; 
	this is a list of lists in the case of multivariate responses (see 
	details below).
\end{enumerate} 
Each of these are addressed briefly below.

@@

% transition and initial state probs
By default, each row of the transition matrix and the vector of
initial state probabilities is modeled as a baseline-category
multinomial logistic model \cite<see>[chapter 7]{Agresti2002}.  This
means that covariates can be used as predictors in these models.  In
particular, the model for each multinomial is:
\begin{equation} 
	p_{i} = \frac{\exp(\alpha_{i}+\mathbf{\beta}^{`}_{i}\vc{x})}
				{1+\sum_{j=1}^{J-1}\exp(\alpha_{j}+\mathbf{\beta}^{`}_{j}\vc{x})},
\end{equation}
where $p_{i}$ is the probability (e.g.\ the probability of the
$i$-th initial state); the $\alpha_{i}$ are the category intercepts;
the $\mathbf{\beta}_{i}$ are the category regression coefficients;
$\vc{x}$ is a vector of covariates or predictors; $J$ is the number of
states; in this example, $J$ serves as the baseline-category, meaning
that $\alpha_{J}$ and $\mathbf{\beta}_{J}$ are zero.

@@

% response models
Response models in \pkg{depmixS4} interface with the \code{glm}
functions \citep{R2008} available in R and with the \pkg{nnet}-package
\citep{Venables2002} for the multinomial logistic models.  Normally
distributed data can hence be modeled with direct effects included as
well.  For example, normal data are modeled as:
\begin{equation}
	O_{t}|S_{t} = \mu + \mathbf{\beta}^{`}\vc{x} + \epsilon_{t},
\end{equation}
where $\mu$ is the (state-dependent) mean; $\vc{x}$ is a vector of
covariates or predictors; $\mathbf{\beta}$ is the vector of
(state-dependent) regression coefficients; $\epsilon_{t}$ is a
normally distributed error term with (state-dependent) standard
deviation $\sigma$.

@@

Two illustrations are provided below of models used to analyze single
participant time series data from two common experimental paradigms.
In both of these, participants learn different rules through trial and
error.  The main goal of these illustrations is to establish the
possibility of analyzing single participant data using the
\pkg{depmixS4}-framework.

@@

% strategic reanalysis of the HDT: different strategies found by
% finite mixture analysis
A reanalysis of this data set \cite{Huizenga2007b} indicated that
participants might solve the task by sequentially considering the
three dimensions (constant gain, frequency of loss, and amount of
loss) in order to choose a door.  Most of the youngest children seem
to focus on frequency of loss, resulting in equal preference for doors
B and D. It is assumed that frequency of loss is the dominant
dimension in this task.  Older participants seem to use a
two-dimensional rule where participants first focus on the frequency
of loss and then consider amount of loss, resulting in a preference
for door D. A third very small subgroup seems to use an integrative
rule where participants combine all three dimensions in the
appropriate way.  Participants using the integrative rule pick cards
from doors C and D, which are both equally advantageous in the long
run, although door D shows a higher volatility.

@@

% problematic aspects of standard analysis
\citet{Huizenga2007b} used the last 60 trials in a series of 200
trials, as is typically done in analyzing this task.  Doing so
silently assumes that behavior has stabilized after 140 trials of
learning; this may not be the case, and it is highly likely that there
are individual differences in this learning process.  Hence, here we
analyze IGT learning data from a single participant with the aim to
establish 1) at which trial learning stops, i.e., at which trial
behavior has stabilized, and 2) what rules participants use
during and at the end of learning.  Models with an increasing number
(from 2 through 5) of latent states were fitted to the time series of
a single participant.  The responses were fitted with multinomial
logistic models for the four possible choices that participants make
in this task.  There were no covariates on any of the parameters.  The
only constraint imposed on the models' parameters was that there were
designated begin and end states.  This means that the initial state
probabilities were fixed to one for the first state and to zero for
the remaining states.  For the transition matrix this means that the
final state was an absorbing state: the probability of transitioning
out of the final state was set to zero.  This was done to ensure that
there would be a final state in which participants end which provides
the possibility of immediately seeing participants' behavior at the
end of the task.  Models were selected using the Akaike Information
Criterion \citep<AIC;>{Akaike1973}.

The data were best described by a 4-state model.  This model's
transition matrix is:
$$
\mat{A} = \begin{pmatrix} 
				0.64 & 0.33 & 0.00 & 0.03 \\
				0.10 & 0.89 & 0.01 & 0.00 \\
				0.05 & 0.00 & 0.95 & 0.00 \\
				0.00 & 0.00 & 0.00 & 1.00
		  \end{pmatrix}
$$
As can be seen from the transition matrix, the final state is
absorbing.  States 2 and 3 are also fairly stable with high diagonal
probabilities of 0.89 and 0.95 respectively.  The response parameters
are presented in Table~\ref{tab:igt4}.

@@

The initial preference for B and D choices confirms the theory
expressed in \citet{Huizenga2007b} that frequency of loss is the
dominant dimension in the IGT. The final states with mostly C 
choices represents one of the optimal rules, which consists of
both C and D responses. Note that C and D responses generate equal
profits in the long run.

@@

There are different accounts of probabilistic category learning
\citep<see e.g.>[for an overview]{Ashby2005}.  According to instance
or exemplar learning theories, participants learn by storing each
encountered cue-outcome pair.  When presented with a probe cue
pattern, a response is made by retrieving these exemplars from memory
and weighting them according to their similarity to the probe cue
pattern.  In contrast, according to associative learning theories,
participants learn by gradually associating the individual cues (or
cue patterns in configural learning) to the outcomes.  Finally, in
rule-learning, participants are taken to extract rules by which to
categorize the different cue patterns.  \citet{Gluck2002} proposed a
number of such rules.  The rules differ in the way the
cues are combined into a response.  The main difference concerns
whether responses are based on the presence/absence of a single cue,
or whether information from multiple cues is integrated.
\citet{Gluck2002} formulated all rules in a deterministic and
optimal manner (e.g., the multi-cue rule corresponded to giving
the optimal response to each cue pattern).  \citet{Meeter2006} allowed
for probabilistic responding (a small probability of responding
non-optimally) as well as switches between rules during learning.
However, both rules and response probabilities remained
predefined.  Alternative non-rule based analyses of the WPT
\cite{Lagnado2006,Speekenbrink2008} have estimated response
distributions through logistic regression, allowing the regression
coefficients to change smoothly over time.

%associative, rule-based.
Here, we combine the regression and rule-based approaches, and
analyze the behavior of a single individual performing the WPT for 200
trials.  We are particularly interested in evidence for rule
switching and whether a DMM can recover rules in accordance with
\citet{Gluck2002}.  We chose to analyze the ``average'' participant
(the participant with performance closest to the group average) in a
large unpublished dataset.  We let each state be characterized by a
GLM with a Binomial distributed response and logistic link function
(i.e., a logistic regression model, see equation 1).  The regression
coefficients of a model relating the cues to the state of the weather
are given in column ```validity'' of Table~\ref{tab:WPT}.  Note that
identical coefficients for the model relating the cues to responses
would indicate probability matching.  A maximizing strategy is
indicated by more extreme regression coefficients.

@@

Investigation of the parameter estimates (see Table~\ref{tab:WPT})
showed that the regression coefficient for the first cue was of much
larger magnitude than that of the other cues.  Because of this, the
first state seems representative of single cue rule.
Alternatively, as all regression coefficients are positive, it could
indicate a ``counting'' heuristic, where the propensity of ``sun''
responses increases when more cues are present (regardless of which
cues they are).  However, in that case, we would expect the regression
coefficients to be of roughly identical magnitude.  The second state
represents a multi-cue rule, as all cues had regression
coefficients of reasonable magnitude in the direction of the objective
cue validities.

To reduce the degrees of freedom and improve parameter estimates, we
implemented constraints to force state 1 into a single cue rule
(fixing the coefficients of the remaining three cues to 0) and state 2
in a multi-cue rule (forcing the intercept to 0).  These
restrictions resulted in a better AIC value of AIC=185.24 (df=7).
Interestingly, the single cue rule was somewhat different than
described by \citet{Gluck2002}.  Parameter estimates indicated relatively
more consistent predictions of ``rain'' in the absence of cue 1
($Pr(\text{sun}) = 0.22$) and more inconsistent predictions of ``sun''
in the presence of cue 1 ($Pr(\text{sun}) = 0.60$).  The cue weights
of the multi-cue rule were in the direction of the optimal
weights.  The Viterbi state sequence indicated that the participant
used the single cue rule for the first 60 trials, and then
switched to the multi-cue rule.

@@

Discrete change is hypothesized to occur in many domains in
experimental psychology and especially so in learning and
developmental tasks.  Piagetian theory predicts discrete changes to
occur in a number of developmental tasks \citep{Inhelder1958}.  There
is a growing body of evidence showing that indeed discrete change
occurs during development and learning
\citep{Jansen2001b,Maas1992,Raijmakers2001,Schmittmann2006}.  Discrete
rules have recently been hypothesized to play an important role in
learning tasks such as the Iowa Gambling Task \citep{Huizenga2007b}
and the Weather Prediction Task \citep{Gluck2002}.  Until now, such
data were always analyzed on a group basis, i.e., pooling data of many
participants together.  Given that it is extremely likely that there
are individual differences between participants, such pooling of data
may not be warranted and may lead to sub optimal results
\citep{Gallistel2004,Molenaar2005}.

In the current study we have shown the feasibility of analyzing single
participant data in two common experimental tasks, thereby avoiding
the pitfalls of pooling data from individuals that may differ in
important respects.  The illustrations show that single participant
data only comprising 200 trials can be analyzed and lead to
non-trivial models with multiple states.  This opens up possibilities
for analyzing data from other experimental tasks on an individual
basis.

In addition to analyzing single participant data, the \pkg{depmixS4}
package \citep{Visser2008b} that we have presented here offers the
possibility of further combining models of single participant data
into larger models and testing whether, for example, some rules are
identical between (groups of) participants.  Performing such analyses
and aggregating models afterwards based on commonalities and
differences between participants avoids making unwarranted assumptions
about homogeneity between participants.  A similar approach using
continuous latent variable models, i.e. factor models, was used by
\citet{Hamaker2007} to integrate within and between person variability
in the domain of personality research.

% IGT
The illustration with the IGT data lead to a four state model.  The
states of the model corresponded very well with earlier analyses by
\citet{Huizenga2007b}, showing a number of discrete rules.  Even
within a single participant, a number of different rules were
observed during the acquisition process; learning started with
behavior close to guessing, which is best interpreted as exploring
behavior, and was followed by two states with close to optimal
performance.  In addition to confirming the existence of discrete
rules, the use of DMM analyses provides us with information of
the dynamics of knowledge acquisition in this task.

% WPT
Analyzing the WPT data on an individual level allowed us to precisely
estimate idiographic rules and their progression during learning.
Moreover, using DMMs, we can combine an individual and group level
analysis, increasing reliability of the individual analyses whilst
allowing for substantial individual differences.  Previous analyses
\cite<e.g.,>{Meeter2006} used pre-defined individual rules, based on
participants' self-reports in \citet{Gluck2002}.  Interestingly,
\citet{Gluck2002} noted that these self-reports often did not
correspond to the rules evident in participants' responses.  It is
well known that people may find it difficult to verbalize the way in
which they integrate multiple sources of information.  This problem is
even more severe in studying young children.  Because of this,
estimating rules directly from participants' responses results in a
more valid assessment of their response rules, and provides a direct
test of the validity of the rule set identified by \citet{Gluck2002}.

% DMM: their great future, conclusion
In sum, DMMs have been shown to be feasible models of individual time
series.  Moreover, \pkg{depmixS4} can be used to build models of
groups of participants combined and allows for testing whether pooling
individual participants' data is warranted.  Due to the use of S4
classes, the \pkg{depmixS4} package is highly flexible and easily
extensible.  Future options are to include other measurement models
such as the factor model or models with autoregressive coefficients.
Also, extensions of the transition model are easily implemented; for
example, the addition of support for continuous time measurement in
contrast to equally space measurement occasions as were used in the
presented illustrations.  Dependent mixture models hence have great
potential in enlarging our insights into important issues in learning
and development, in particular in modeling and testing the occurrence
of discrete change.
